The E-C Algorithm: conceptual bootstrapping via grammar-based program compression.

### 2013-01-20 drafts/ijcai2013/ijcai13.tex
Changes.

As a simple example of our representation, consider the squaring
function $\lamba x. * x x $. Using two of the basis combinators above,
we can write the squaring function in combinatory logic as $S * I$,
where application associates to the left, so we drop the
parentheses. When we apply this combinator to a value $x$, the action
of the combinator is defined as $S \;*\; I\; x~\rightarrow~(*\; x)\;
(I\; x)~\rightarrow(*\; x)\; x$. Now, we can extend this
representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes are typed applications of combinators.

\begin{figure}
  \centering
  \begin{subfigure}{0.2\linewidth}
    \label{fig:clbintree}
    \begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}]
      \node [font=\scriptsize] (top) {$I$}
      child {node [above=.3cm] (a) {$(I \rightarrow I) \rightarrow I \rightarrow I$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$I \rightarrow I$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };

    \end{tikzpicture}
  \end{subfigure}

  \begin{subfigure}[b]{0.2\linewidth}
    \label{fig:andor}
    \begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1}]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {}
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$}}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$}} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

      \path (d)  edge node[inner sep=0mm,pos=.4] (a1) {} node {} (l);
      \path (d)  edge node[inner sep=0mm,pos=.4] (b1) {} node {} (r);
      \path (a1) edge[bend right] (b1) ;

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}

    \end{subfigure}
  \caption{A typed binary tree representation of the combinator $S\;*\;I\;$.}
\end{figure}

\subsection{Best-first enumeration of most probably programs.}
In order to explore the frontier of most probable programs, we need a
procedure that enumerates the $N$ most probable programs. We can
formulate this naturally as a best-first exploration of an $AND-OR$
tree (NEED FIGURE). Complete programs are represented by binary trees
that are valid subtrees of the $AND-OR$ graph. The tree structure is
shown in Figure~\ref{fig:andor}. The root is an $OR$ node of the
requisite program type. It's children are the primitives of that type
and one $AND$ node with two children. Each child has the same
structure as the root node, except that the types are different. If
the type of an $OR$ node is $t$, then the type of the left child of
its $AND$ node child is $\tau \rightarrow t$ (where $\tau$ is a fresh
type variable). Since the type of the right child is constrained by
the subtree rooted at the left child, we always expand the left child
first, and we use the target type of this left subtree to mark the
type of the right child.

\begin{figure}
\label{fig:andor}
\begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1}]
\node [circle,draw] (a) {}
child {node [left=0cm] (b) {$f_1$}}
child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
child {node [left=.4cm] (e) {$f_n$}}
child {node [circle,draw, left=1cm] (d) {}
  child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
    child {node [left=.5cm] (b) {$f_1$}} 
    child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
    child {node [left=.3cm] (e) {$f_n$}}
    child {node [circle, draw, left=.25cm] (and2) {}
          child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
          child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
    }
  }
  child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
    child {node [circle, draw, right=.25cm] (and3) {}
          child {node [circle, draw ](l3) {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
          child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
    }
    child {node [right=.3cm] (b) {$f_1$}} 
    child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
    child {node [right=.6cm] (e) {$f_n$}}
 }
};

\path (d)  edge node[inner sep=0mm,pos=.4] (a1) {} node {} (l);
\path (d)  edge node[inner sep=0mm,pos=.4] (b1) {} node {} (r);
\path (a1) edge[bend right] (b1) ;

\path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
\path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
\path (t1) edge[bend right] (k1) ;
\end{tikzpicture}
\end{figure}

@@

Given a set of solution programs, we want to re-estimate our
stochastic grammar. Our algorithm for doing this is based broadly on
the Neville-Manning algorithm for grammar-based
compression~\cite{nevill1997identifying}. In our algorithm, we simply
count the numer of times each unique subtree appears as the child of a
subtree \emph{not yet encountered} in the solution set. That is, we
traverse the solution set in depth first order, comparing each tree we
encounter to all the trees we previously encountered; but if we
encounter a tree we have already seen, we do not descend into it. In
this traversal, the set of trees we encounter more than once is the
set of terminal nodes in our new grammar. To estimate the
probabilities associated with each node production, we again traverse
the solution set, but these time for each node where a terminal
production takes place we count which other terminal productions could
have taken place. We estimate the terminal production weight to be the
ratio of the number of times that node was used in the first traversal
divided by the number of times it was used in the second traversal
(plus 1, to account for the event of an $AND$ node
expansion). Likewise, we re-estimate the probability of an $AND$ node
expansion by diving the number of such expansions in the solution set
by the number of possible expansions in the solution set.

@@

\begin{figure}
\label{fig:symreg}
\includegraphics[width=\linewidth]{figures/learningCurves.pdf}
\includegraphics[width=\linewidth]{figures/performanceFrontierSize.pdf}
\caption{E-C learning curves for the symbolic regression task set as
  we vary the frontier size.}
\end{figure}

>>>>

As a simple example of our representation, consider the squaring
function $\lambda x. * x\; x $. Using two of the basis combinators above,
we can write the squaring function in combinatory logic as $S * I$,
where application associates to the left, so we drop the
parentheses. When we apply this combinator to a value $x$, the action
of the combinator is defined as $S \;*\; I\; x~\rightarrow~(*\; x)\;
(I\; x)~\rightarrow(*\; x)\; x$. Now, we can extend this
representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes are typed applications
of combinators.

\begin{figure}
  \centering
  \begin{subfigure}[Before]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7,
        transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$I$}
      child {node [above=.3cm] (a) {$(I \rightarrow I) \rightarrow I \rightarrow I$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$I \rightarrow I$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
    \caption{The typed combinator $S * I$ ($f(x)=x^2$) represented as binary tree.}
    \label{fig:clbintree}
  \end{subfigure}
  \begin{subfigure}[After]{0.4\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

      \path (d)  edge[very thick]  node[inner sep=0mm,pos=.4] (a1) {} node {} (l);
      \path (d)  edge[very thick]  node[inner sep=0mm,pos=.4] (b1) {} node {} (r);
      \path (a1) edge[bend right] (b1) ;

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \caption{The space of all typed combinators represented as an
      infinitely deep And-Or tree. Any specific combinator as a
      sub-binary tree of this And-Or tree whose leaves consist of only
      the $f_i$ nodes (i.e., the primitive combinators of the
      grammar).}
    \label{fig:andor}
    \end{subfigure}
  \caption{A typed binary tree representation of the combinator $S\;*\;I\;$.}
\end{figure}

\subsection{Best-first enumeration of most probably programs.}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. We can
formulate this naturally as a best-first exploration of an $AND/OR$
tree~\cite{nilsson1982principles}, which we denote $\mathcal{G}$. A
complete program is any binary subtrees of $\mathcal{G}$ whose leaves
coincide with the leaves of $\mathcal{G}$. The tree structure is shown
in Figure~\ref{fig:andor}. The root is an $OR$ node of the requisite
program type. It's children are the primitives of that type $f_1,
\dots, f_n$, and one $AND$ node with two children. Each child of an
$AND$ node has the same structure as the root node, with modified
types: if the type of an $AND$ node is $t$, then the type of its left
child is $\tau \rightarrow t$ (where $\tau$ is a fresh type
variable). Since the type of the right child is constrained by the
subtree rooted at the left child, we always expand the left child
first and use the target type of this left subtree to mark the type of
the right child.

@@

Given a set of solution programs, we want to re-estimate our
stochastic grammar. Our inspiration for doing this is based broadly on
the Neville-Manning algorithm for grammar-based
compression~\cite{nevill1997identifying}. The idea of that compression
algorithm is to compress any string by creating a grammar for that
string such that a) no two symbols appear more than once in the
grammar and b) every rule is used more than once. From a set of trees
we generate a grammar according to these criteria (though using a less
efficient and much simpler algorithm).

In our algorithm, we traverse a set of combinator binary trees in
depth first order. We count the occurences of every subtree. Every
time we encounter a subtree for the second time, we designate it as a
primitive in our new grammar. Furthermore, during our traversal, we do
not descend into trees that have been added already to the new
grammar. Thus, we only count an instance of tree if that instance is
not a subtree of a primitive element of the new grammar.

We want to count the number of
times every subtree is encountered when it is not a subtree of a tree
already encountered in our traversal. we simply count the numer of
times each unique subtree appears as the child of a subtree \emph{not
  yet encountered} in the solution set. That is, we traverse the
solution set in depth first order, comparing each tree we encounter to
all the trees we previously encountered; but if we encounter a tree we
have already seen, we do not descend into it. In this traversal, the
set of trees we encounter more than once is the set of terminal nodes
in our new grammar. To estimate the probabilities associated with each
node production, we again traverse the solution set, but these time
for each node where a terminal production takes place we count which
other terminal productions could have taken place. We estimate the
terminal production weight to be the ratio of the number of times that
node was used in the first traversal divided by the number of times it
was used in the second traversal (plus 1, to account for the event of
an $AND$ node expansion). Likewise, we re-estimate the probability of
an $AND$ node expansion by diving the number of such expansions in the
solution set by the number of possible expansions in the solution set.

@@

\begin{figure}

\includegraphics[width=\linewidth]{figures/learningCurves.pdf}
\includegraphics[width=\linewidth]{figures/performanceFrontierSize.pdf}
\caption{E-C learning curves for the symbolic regression task set as
  we vary the frontier size.}
\label{fig:symreg} 
\end{figure}
